---
title: Reanalysis of the Care Management by Oncology Nurses to Address Supportive Care Needs cluster-randomized trials
toc: True
output: pdf_document
fig_caption: yes
bibliography: ref.bib 
---

We applied our developed method [@sace] to estimate the survivor average causal effect (SACE) as an evaluation of the intervention effect in the  clustered randomized clinical trial of the CONNECT (Care Management by Oncology Nurses to
Address Supportive Care Needs) study [@connect]. 
The trial  was conducted from July 25, 2016, to October 6, 2020 at 17 community general medical
oncology practices within the University of Pittsburgh Medical
Center Hillman Cancer Center Network in western Pennsylvania. Participants of the study were adult patients (aged $\geq 21$ years) with metastatic
solid tumors for whom the oncologist would agree
with the statement ``would not be surprised if the patient
died in the next year." CONNECT used an oncology
nurse led care management approach to improve the provision
of primary palliative care within outpatient oncology
practices. The clinical team's infusion room nurses were trained to address deficient
care processes within their oncology practices.
The intervention was designed to occur monthly over a
3-month period, with CONNECT visits taking place before and/or
after regularly scheduled oncology clinic visits. Enrolled patients
who were randomized to CONNECT met with the same
nurse for all visits, and nurses had the option of conducting these
visits by telephone. Enrolled patients who were randomized to standard care received
oncology care according to best practices, including all
supportive measures deemed appropriate by the oncology
team.

```{r setup, echo = F,message=F,warning=F}
knitr::opts_chunk$set(echo = F,message=F,warning=F)
library(knitr)
da=read.csv("CONNECT_Primary_Paper_Final.csv")
hm=model.matrix(~as.factor(da$PT_ecog)-1)
colnames(hm)=paste0("ecog",0:2)
da=cbind(da,hm)
da$treat=(da$random_id==1)*1
death="PatientDied_Before_3M"
ocs=c("PT_3M_FACIT_Pal_Score","PT_3M_ESAS_Score",
"PT_3M_HADS_Dep_Score","PT_3M_HADS_Anx_Score")
ocb=c("PT_BL_FACIT_Pal_Score","PT_BL_ESAS_Score",
"PT_BL_HADS_Dep_Score","PT_BL_HADS_Anx_Score")
sda=da[,c(death,"treat","PT_age","PT_BL_chemotherapy",ocs,ocb,paste0("ecog",0:2),"hospital_id")]
tt=table(sda$treat)
td=table(sda[,death],sda$treat)
```

Participants completed questionnaires at baseline and 3
months, either through a telephone interview or on paper.
Demographic data were collected from baseline questionnaires.
The primary outcome was quality of life. At baseline and 3
months, participants completed assessments of quality of life, FACIT  (Functional Assessment of
Chronic Illness Therapy-Palliative care) score. 
The score ranges from 0 to 184, with higher scores indicating
better quality of life. Secondary outcomes include 
the Edmonton Symptom Assessment
Scale (ESAS) score which ranges from 0 to 90, with
higher scores indicating greater symptom burden.
The hospital anxiety subscale score and the hospital depression
subscale score were also used where each score ranges from 0
to 21 and
scores $\geq 8$ indicates substantial anxiety or depression
symptoms. However, the outcome scores were not measured if participants died, withdrew from the study, or were lost to follow-up. 

The data set in our analysis has `r nrow(sda)` patients, 
`r tt[2]` (`r round(100*tt[2]/sum(tt),2)`\%) are in the intervention arm. 
During the study period, `r td[2,2]` (`r round(100*td[2,2]/sum(td[,2]),2)`\%) patients died in the intervention arm and `r td[2,1]` (`r round(100*td[2,1]/sum(td[,1]),2)`\%) patients died in the standard care arm. 
 Table 1 shows the truncation frequency and the percentage for each
of the outcome scores in each arm due to the reasons above. 
Similar truncation percentages are observed for all the outcomes. 

```{r}
locs=c("FACIT-Pal total score","ESAS total score","Depression subscale score","Anxiety subscale score")
tcount=lapply(sda[,ocs],function(x){table(sda$treat,is.na(x))[,2]})
tc=matrix(unlist(tcount),4,2,byrow=T)
tprop=lapply(sda[,ocs],function(x){round(prop.table(table(sda$treat,is.na(x)),1)[,2],4)})
tp=matrix(unlist(tprop),4,2,byrow=T)*100
tcp=cbind(paste0(tc[,1],", ",tp[,1]),paste0(tc[,2],", ",tp[,2]))
rn=c("FACIT","ESAS","Depression","Anxiety")
```

```{r}
rownames(tcp)=rn
kable(tcp,'simple',caption="Truncation frequency and percentage for each outcome score by the
arm.",col.names=c("CONNECT arm","Standard care arm"))
```

Figure 1 further shows the truncation proportion of the primary outcome FACIT score by cluster size
for each arm. The plot suggests that the proportion is independent of the arm and also the cluster size. 
Similar patterns are observed for the other outcome scores and are thus not shown. 

```{r, fig.cap="The proportion of truncation for the outcome FACIT score in each cluster vs. the cluster size by arm"}
tr.name="treat"
cluster.name="hospital_id"
sind=((is.na(sda[,ocs[1]]))*1)
ts=table(sind,sda[,cluster.name],sda[,tr.name])
tc=ts[,,1]
otcs=colSums(tc)
tcs=otcs[which(otcs>0)]
ptc=tc[2,which(otcs>0)]/tcs
tt=ts[,,2]
otts=colSums(tt)
tts=otts[which(otts>0)]
ptt=tt[2,which(otts>0)]/tts
par(mfrow=c(1,2),mar = c(4,0,3,0), oma = c(3,3,1,3),cex.axis=1.2)
plot(tts,ptt,ylim=c(0,1),las=1,xlab="CONNECT arm",ylab="Proportion of truncation")
plot(tcs,ptc,yaxt="n",ylim=c(0,1),xlab="Standard care arm")
axis(4,las=1)
mtext("Cluster size",1,outer=T,cex=1,line=-0.2)
mtext(locs[1],3,outer=T,line=-1.5,cex=1.3)
```

We apply the proposed approaches in [@sace]
to analyze each score at 3 months adjusting for the corresponding baseline score, age, chemotherapy
treatment, and the Eastern Cooperative Oncology Group Performance Status score (0, 1 or 2). 
For ease of reference, we abbreviate
the fixed-effects only approach there as the FE approach, the approach with clustering in the outcome model only as the
ME approach, and the approach with clustering in the outcome model and the membership model as the ME2 approach.
R functions to implement the methods (FE, ME and ME2) are available online at 
https://github.com/harhay-lab/SACE\_PS\_LMM.

```{r}
library(doParallel)
tr.name="treat"
cluster.name="hospital_id"
van=c("PT_age","PT_BL_chemotherapy",paste0("ecog",1:2))
ind=which(rowSums(is.na(sda[,c(van,ocb)]))==0)
ssda=sda[ind,]
rm(sda)
source("fecode.R")
source("mecode.R")
source("me2code.R")
source("bfun.R")
uc=sort(unique(ssda[,cluster.name]))
luc=length(uc)
tc=table(ssda[,cluster.name],ssda[,tr.name])
n=nrow(ssda)
sda1=subset(ssda,treat==1)
tcid=sort(unique(sda1[,cluster.name]))
ltcid=length(tcid)
n1=nrow(sda1)
sda0=subset(ssda,treat==0)
ccid=sort(unique(sda0[,cluster.name]))
lccid=length(ccid)
n0=nrow(sda0)
sacefit=function(y.name,yb.name){
ff=fsfun(ssda,y.name,tr.name,c(van,yb.name))
cf=csfun(ssda,y.name,tr.name,c(van,yb.name),cluster.name)
rf=rcsfun(ssda,y.name,tr.name,c(van,yb.name),cluster.name)
return(list(ff=ff,cf=cf,rf=rf))}
sfit=vector("list",4)
for(k in 1:4){sfit[[k]]=sacefit(ocs[k],ocb[k])}
icc=round(rbind(c(sfit[[1]]$cf[2],sfit[[1]]$rf[2],sfit[[1]]$rf[3]),
c(sfit[[2]]$cf[2],sfit[[2]]$rf[2],sfit[[2]]$rf[3]),
c(sfit[[3]]$cf[2],sfit[[3]]$rf[2],sfit[[3]]$rf[3]),
c(sfit[[4]]$cf[2],sfit[[4]]$rf[2],sfit[[4]]$rf[3])),2)
rownames(icc)=rn
kable(icc,'simple',caption="The estimated outcome ICC from the ME and ME2 approaches,
and the estimated ICC of the strata membership by the ME2 approach.",
col.names=c("Outcome model (ME)","Outcome model (ME2)","Membership model (ME2)"))
```

The estimated outcome model intracluster correlation coefficients (ICC's) are in Table 2 from the ME and ME2 approaches,
and the estimated ICC of the strata membership by the ME2 approach. The outcome model ICC's are similar by the two approaches. 

```{r}
esace=round(rbind(c(sfit[[1]]$ff[1],sfit[[1]]$cf[1],sfit[[1]]$rf[1]),
c(sfit[[2]]$ff[1],sfit[[2]]$cf[1],sfit[[2]]$rf[1]),
c(sfit[[3]]$ff[1],sfit[[3]]$cf[1],sfit[[3]]$rf[1]),
c(sfit[[4]]$ff[1],sfit[[4]]$cf[1],sfit[[4]]$rf[1])),2)
rownames(esace)=rn
```

```{r}
ncc=detectCores()
nbp=10
pb=c(0.025,0.975)
bci=matrix(0,4,6)
for(j in 1:4){
y.name=ocs[j]
xnames=c(van,ocb[j])
cl=makeCluster(ncc)
registerDoParallel(cl)
fsobj=foreach(icount(nbp),.combine='rbind',.multicombine=T,.maxcombine=nbp,
.errorhandling ="remove") %dopar% {
ind1=sample(n1,n1,replace=T)
ind0=sample(n0,n0,replace=T)
fsfun(rbind(sda1[ind1,],sda0[ind0,]),y.name,tr.name,xnames)}

csobj=foreach(icount(nbp),.combine='rbind',.multicombine=T,.maxcombine=nbp,
.errorhandling ="remove") %dopar% {
source("mecode.R")
source("bfun.R")
s1=sample(tcid,ltcid,replace=T)
s0=sample(ccid,lccid,replace=T)
bda1=bda0=NULL
for(jj in 1:ltcid){
da1j=subset(sda1,hospital_id==s1[jj])
da1j$bcid=jj
bda1=rbind(bda1,da1j)}
for(kk in 1:lccid){
da0k=subset(sda0,hospital_id==s0[kk])
da0k$bcid=kk
bda0=rbind(bda0,da0k)}
csfun(rbind(bda1,bda0),y.name,tr.name,xnames,"bcid")}

rsobj=foreach(icount(nbp),.combine='rbind',.multicombine=T,.maxcombine=nbp,
.errorhandling ="remove") %dopar% {
source("me2code.R")
source("bfun.R")
s1=sample(tcid,ltcid,replace=T)
s0=sample(ccid,lccid,replace=T)
bda1=bda0=NULL
for(jj in 1:ltcid){
da1j=subset(sda1,hospital_id==s1[jj])
da1j$bcid=jj
bda1=rbind(bda1,da1j)}
for(kk in 1:lccid){
da0k=subset(sda0,hospital_id==s0[kk])
da0k$bcid=kk
bda0=rbind(bda0,da0k)}
rcsfun(rbind(bda1,bda0),y.name,tr.name,xnames,"bcid")}
stopCluster(cl) 
bci[j,]=c(quantile(fsobj[,1],prob=pb),
quantile(csobj[,1],prob=pb),quantile(rsobj[,1],prob=pb))}
bci=round(bci,2)
bsace=cbind(paste0(esace[,1]," (",bci[,1],", ",bci[,2],")"),
paste0(esace[,2]," (",bci[,3],", ",bci[,4],")"),
paste0(esace[,3]," (",bci[,5],", ",bci[,6],")"))
rownames(bsace)=rn
```

Table 3 shows the SACE estimates together with their $95\%$ confidence intervals based on `r nbp` bootstrap replicates. 
We obtain the confidence interval of the estimate using the bootstrap method of sampling the clusters with replacement.
We randomly sample the treated clusters and the control clusters separately. For each bootstrap sample, we fit the
model and obtain the SACE estimate. The $95\%$ confidence intervals are obtained from the $2.5\%$ and the $97.5\%$ quantiles of the
bootstrap estimates.
```{r}
kable(bsace,'simple',caption="The estimated SACE by the FE, ME and ME2 approaches.",
col.names=c("FE","ME","ME2"))
```
